Predicting computational reproducibility of scientific pipelines using collaborative filtering

### 2018-07-23 paper.tex
Fixed normalization in distributions. Worked on results. Started working on intro.

We consider the problem of predicting the values of a variable 
controlled by a hidden, parameterized, time dependent, deterministic 
process. We assume that the variable values are observed only for 
a limited number of time points, and we aim at predicting the values in the remaining time points.

We approach the problem from the point of view of collaborative 
filtering. The process is represented by a matrix where columns 
represent different parametrizations (we do not assume an order on the 
parametrizations) and rows represent the time points. The main difference with the traditional 
collaborative filtering problem

% This is just a working example

Computational reproducibility, the property through which
computational results can be recomputed over time and
space~\cite{stodden}, has become a critical component of scientific
methodology with the rise of the reproducibility crisis in several
domains~\cite{xxx}. Among the factors hampering computational
reproducibility, infrastructural characteristics such as the operating
system play an important role. In neurosciences, our primary field of
interest, several studies have shown the effect of the operating
system on computational results. However, conducting such
reproducibility studies at scale is cumbersome due to the execution
time of data analysis pipelines, which easily exceeds a few hours.

@@

% Computational reproducibility is an issue, for instance among
% different operating systems (refer to Glatard FINF 2015,
% Gronenschild 2012).

@@

% Refer to Germain et al 2008.

@@

% Problem: predict the reproducibility of pipeline files from other
% subjects and the first files produced by a pipeline. Restrict the
% study to binary classification.

@@

\todo{Cite Cecile's paper}

@@

@@

@@

          \textit{a} = (2\alpha - 1) N_{f} & \text{if $\alpha > 0.5$}
  \end{cases}
\]
This method is realistic as files are sampled according to their production time-stamps.
Due to sampling issues, it is possible that the actual training ratio obtained with this method
does not match the target one. We check that the difference between the target and real training ratios
was lower than 0.01.\\

\paragraph{Random File Numbers (Triangular) -- RFNT}
The number of files selected for every subject is randomly selected in
a triangular distribution with parameters \textit{a}, \textit{b} and 
\textit{c} as in Figure~\ref{fig:triangular}. The mean of the distribution is 
$\frac{a+b+c}{3}$. We set \textit{c} to $N_{f}$ and we set 
\textit{a} and \textit{b} with two approaches that ensure that the 
average of the distribution is $\alpha N_{f}$ ($\alpha$ is the training 
ratio):
\begin{enumerate}
        \item Largest a (RFNT-L, Figure~\ref{fig:triangular-L-Sample-Training-set}): a is set to the 
        largest possible value, i.e., b. The average of the 
        distribution is $\frac{2a+N_{f}}{3}$, therefore 
        $a=\frac{3\alpha-1}{2}N_{f}$.
                \item Smallest a (RFNT-S, Figure~\ref{fig:triangular-S-Sample-Training-set}): a is 
        set to the smallest possible value, i.e., 0. The average of the 
        distribution is $\frac{b+N_{f}}{3}$, therefore 
        $b=N_{f}(3\alpha-1)$. To ensure that $b<N_{f}$, we actually set 
        $b$ to $\min(N_{f}, N_{f}(3\alpha-1))$. 
\end{enumerate}
It should be noted that $\alpha$ must be larger than $\frac{1}{3}$ for 
a and b to remain positive. For training ratios lower than $\frac{1}{3}$, we 
define RFNT methods as the RFNU one presented previously.
As illustrated in Figures~\ref{fig:triangular-L-Sample-Training-set} 
and~\ref{fig:triangular-S-Sample-Training-set}, the motivation for the 
RFNT-L method is to guarantee that, for large enough values of 
$\alpha$, all subjects will have at least a few files in the training 
set, which is not the case for RFNU.
\begin{figure}
\centering
\includegraphics[width=0.5\columnwidth]{figures/triangular.pdf}
\caption{Triangular distribution}
\label{fig:triangular}
\end{figure}

@@

We compare our sampling methods to (1) a dummy classifier that always predicts the value in the majority class
 and (2) the Random Unreal method, used as the baseline sampling technique.

@@

Results of ALS with Bias are reported in Figure~\ref{fig:results-synthetic-als-bias}.
RS, RFNU and RFNTL remain the methods that best compare to Random Unreal, but the achieved
accuracy is much lower than without Biases. This is due to the fact that biases \ldots \todo{finish that.}

@@

@@

\begin{figure*}
\begin{subfigure}[b]{0.4\linewidth}
        \includegraphics[width=0.8\columnwidth, angle=-90]{data/results/means_of_results/ALS-Bias/PreFreeSurfer/ALS-Bias-PFS-5v6.pdf}
        \caption{PFS, CentOS5 vs CentOS6}
\end{subfigure}
\begin{subfigure}[b]{0.4\linewidth}
        \includegraphics[width=0.8\columnwidth, angle=-90]{data/results/means_of_results/ALS-Bias/PreFreeSurfer/ALS-Bias-PFS-5v7.pdf}
        \caption{PFS, CentOS5 vs CentOS7}
\end{subfigure}
\begin{subfigure}[b]{0.4\linewidth}
        \includegraphics[width=0.8\columnwidth, angle=-90]{data/results/means_of_results/ALS-Bias/PreFreeSurfer/ALS-Bias-PFS-6v7.pdf}
        \caption{PFS, CentOS6 vs CentOS7}
\end{subfigure}\hfill
\begin{subfigure}[b]{0.4\linewidth}
        \includegraphics[width=0.8\columnwidth, angle=-90]{data/results/means_of_results/ALS-Bias/FS-100files/ALS-Bias-FS100files.pdf}
        \caption{FS, CentOS6 vs CentOS7}
\end{subfigure}
\caption{Accuracy results on PreFresurfer (PFS) and Freesurfer (FS) data, ALS \emph{with} bias.}
\label{fig:results-real-als-bias}
\end{figure*}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% General conclution on results of ALS without Bias 
%~ In general the accuracy pattern of the two Complete Rows and 
%~ RS methods show similar decline and grow behaviour over the 
%~ training ratio axis however it can be subject to considerable 
%~ fluctuations in differentiated condition pairs. Despite of their 
%~ likely irrational behaviour, they always have better accuracy 
%~ for the training size with less than $0.3$ in comparison to other 
%~ methods. This better performance (around 75 percent)is almost three 
%~ times higher than other methods in condition pairs with more 
%~ differences but just 20 percent better when the conditions are more 
%~ similar to each other (CentOS5 vs CentOS6) with accuracy of $90\%$.
%~ Figure~\ref{fig:RS_Rows} represents
%~ the accuracy graph of these two methods with comparison to dummy
%~ classifier over the increment of the training ratio.
%~ \begin{figure}[h!]
        %~ \centering
        %~ \begin{subfigure}[b]{0.4\linewidth}
                %~ \includegraphics[width=\columnwidth]{figures/RS_Rows_5vs6}
        %~ \end{subfigure}
        %~ \begin{subfigure}[b]{0.4\linewidth}
                %~ \includegraphics[width=\columnwidth]{figures/RS_Rows_5vs7}
        %~ \end{subfigure}
        %~ \begin{subfigure}[b]{0.4\linewidth}
                %~ \includegraphics[width=\columnwidth]{figures/RS_Rows_6vs7}
        %~ \end{subfigure}
        %~ \caption{Average performance of RS and Complete Rows methods in three different condition pairs}
        %~ \label{fig:RS_RowS}
%~ \end{figure}

%~ All the other random methods, RS and both RFNT approaches 
%~ are performing a continues growth Figure~\ref{fig:simple-methods} represents
%~ the accuracy graph of all random methods with comparison to dummy
%~ classifier over the increment of the training ratio.. Although some considerable variation 
%~ for both triangular approaches have been observed (CentOS5 vs CentOS6) 
%~ their constant rise leads us to the next phase of experiments which was 
%~ adding up bias into the current ALS technique. 
%~ The training set suffers from lacking the files which are almost generated towards the end of pipeline.

%~ \begin{figure}[h!]
        %~ \centering
        %~ \begin{subfigure}[b]{0.4\linewidth}
                %~ \includegraphics[width=\columnwidth]{figures/simple-methods-5vs6}
        %~ \end{subfigure}
        %~ \begin{subfigure}[b]{0.4\linewidth}
                %~ \includegraphics[width=\columnwidth]{figures/simple-methods-5vs7}
        %~ \end{subfigure}
        %~ \begin{subfigure}[b]{0.4\linewidth}
                %~ \includegraphics[width=\columnwidth]{figures/simple-methods-6vs7}
        %~ \end{subfigure}
        %~ \caption{Average performance of all methods in three different condition pairs}
        %~ \label{fig:simple-methods}
%~ \end{figure}

%-----------------------------------------------------------
% Present your results: accuracy, ROC curves, transparency matrix, factors.

% Try with different numbers of factors.

\section{Discussion}
this is just a test of compiling

% Not all subjects behave the same, which motivates the Big Data approach.

% Which sampling method is best

% Interpreting the factors? Factors reflect the pipeline definition.

% How can this be used in practice

% What are the limitations

% Real values, we expect the best sampling method to remain the same.

\section{Conclusion}

% Summary of the results and discussion. The take-home message.
%%%%%% why there is a drop in ALS-RFNT-L with 60% ratio?

>>>>

Computational reproducibility, the property through which
computational results can be recomputed over time and
space~\cite{peng2011reproducible}, has become a critical component of scientific
methodology with the rise of the reproducibility crisis in several 
domains~\cite{baker2016there}. Among the factors hampering 
computational reproducibility, infrastructural characteristics such as 
the operating system play an important role. In neurosciences, our 
primary field of interest, several studies have shown the effect of the 
operating system on computational 
results~\cite{gronenschild2012effects, glatard2015reproducibility}. 
However, conducting such reproducibility studies at scale is cumbersome 
due to the execution time of data analysis pipelines, which easily 
exceeds a few hours.

We consider the problem of predicting the values of a variable 
controlled by a hidden, parameterized, time-dependent, deterministic 
process. We assume that the variable values are observed only for a 
limited number of time points and process parameters, and we aim at 
predicting the values for the remaining time points and parameters.

This problem description maps directly to the evaluation of 
computational reproducibility in scientific pipelines: the process is 
the pipeline, it is hidden since we consider it as a black box (we 
don't inspect its source code), it is parametrized as it behaves 
differently on different input datasets and parameters, it is 
time-dependent as the value of the variable of interest varies 
throughout the course of the pipeline execution. The variable of 
interest is whether the pipeline outputs, most likely the files it 
produces, are identical between execution conditions.

We approach the problem from the point of view of collaborative 
filtering, inspired by previous examples that applied this method 
outside of its initial application domain~\cite{feng2013efficient}. The 
process is represented by a matrix where columns represent different 
parametrizations and rows represent the files produced by the 
execution. Parmeterizations may represent different input data sets or 
parameter values, on which we do not assume an order. On the contrary, 
an order is defined on the files, from their last modification time in 
a sequential execution. We assume that files are produced in the same 
order regardless of the parametrization, which is not always true, in 
particular when pipelines are multi-threaded, but in general 
reasonable \todo{explain "local" truth of this assumption}. This order imposes a constraint on the construction of the
training set since files

@@

@@

@@

@@

@@

\todo{talk about time-constrained processes}

The paper makes the following contributions:
\begin{enumerate}
\item We model reproducibility evaluations in data processing pipelines as a collaborative filtering problem.
\item We propose strategies to sample the training set under time constraints.
\item We evaluate and compare our sampling strategies on synthetic and real datasets.
\end{enumerate}

@@

\todo{Restrict to binary classification}

@@

          \textit{a} = (2\alpha - 1) N_{f} & \text{if $\alpha > 0.5$}
  \end{cases}
\]
 For $\alpha > 0.5$, this ensures that the average number of files in 
 the training set is $\alpha N_f$. For $\alpha \leq 0.5$, the number of 
 files selected in every subject is sampled from $U(0,N_f)$ with 
 probability $2\alpha$, and set to 0 otherwise.
RFNU is realistic as files are sampled according to their production time-stamps.
Due to sampling issues, it is possible that the actual training ratio obtained with this method
does not match the target one. We check that the difference between the target and real training ratios
was lower than 0.01.\\

\paragraph{Random File Numbers (Triangular) -- RFNT}
The number of files selected for every subject is randomly selected in
a triangular distribution $T(a, b, c)$ as in Figure~\ref{fig:triangular}. The mean of the distribution is 
$\frac{a+b+c}{3}$. We set \textit{c} to $N_{f}$ and we set 
\textit{a} and \textit{b} with two approaches that ensure that the 
average of the distribution is $\alpha N_{f}$ ($\alpha$ is the training 
ratio):
\begin{enumerate}
        \item Largest a (RFNT-L, Figure~\ref{fig:triangular-L-Sample-Training-set}): a is set to the 
        largest possible value, i.e., b. The average of the 
        distribution is $\frac{2a+N_{f}}{3}$, therefore 
        $a=\frac{3\alpha-1}{2}N_{f}$. When $\alpha \leq 1/3$, the 
        number of files in every subject is sampled from $T(0, 0, N_f)$ 
        with probability $3\alpha$ and set to 0 otherwise, to ensure that the average number of files
        in the training set is $\alpha N_f$.
        \item Smallest a (RFNT-S, Figure~\ref{fig:triangular-S-Sample-Training-set}): a is 
        set to the smallest possible value, i.e., 0. The average of the 
        distribution is $\frac{b+N_{f}}{3}$, therefore 
        $b=N_{f}(3\alpha-1)$. This implies that $\alpha \geq 1/3$, so 
        that $b \geq 0$, and $\alpha \leq 2/3$, so that $b \leq N_f$. 
        When $\alpha < 1/3$, we set $b=0$ and the number of files 
        selected in every subject is sampled from $T(0, 0, N_f)$ with 
        probability $3\alpha$ and set to 0 otherwise, as in RFNT-L. 
        When $\alpha \geq 2/3$, we set $b=N_f$ and the number of files 
        selected in every subject is sampled from $T(0, N_f, N_f)$ with 
        probability $3(1-\alpha)$ and set to $N_f$ otherwise, still to 
        ensure that the average number of files in the training set is 
        $\alpha N_f$.
\end{enumerate}
As illustrated in Figures~\ref{fig:triangular-L-Sample-Training-set} 
and~\ref{fig:triangular-S-Sample-Training-set}, the motivation for the 
RFNT-L method is to guarantee that, for large enough values of 
$\alpha$, all subjects will have at least a few files in the training 
set, which is not the case for RFNU.
\begin{figure}
\centering
\includegraphics[width=0.5\columnwidth]{figures/triangular.pdf}
\caption{Triangular distribution $T(a, b, c)$}
\label{fig:triangular}
\end{figure}

@@

We compare our sampling methods to (1) a dummy classifier that always 
predicts the value in the majority class and (2) the Random Unreal 
method, used as the baseline sampling technique. All reported values are averages
over 5 repetitions. \todo{Report model parameters, including number of factors.}

@@

Results of ALS with bias are reported in Figure~\ref{fig:results-synthetic-als-bias}.
RS, RFNU and RFNTL remain the methods that best compare to Random Unreal, but the achieved
accuracy is much lower than without Biases. This is due to the fact that biases \ldots \todo{finish that.}

@@

Results for ALS without bias are reported in 
Figure~\ref{fig:results-real-als}. Among the methods that performed 
well in the synthetic dataset, RFNU performs the best, with an accuracy
higher than 0.95 \todo{check value} for all datasets, when the training ratio is larger than 0.5. RS and RFNTL
are slightly underperforming compared to RFNU.
Complete rows and complete columns do not even reach the accuracy of the 
dummy classifier. The reference method, Random Unreal, has an accuracy close
to 1, which shows that collaborative filtering works well on this dataset too.

@@

Results for ALS with bias are reported in 
Figure~\ref{fig:results-real-als-bias}. From a training ratio of 0.5, 
all methods have a very good accuracy for all datasets, except complete 
columns in Figure~\ref{fig:pfs-c5vsc7-bias}. All methods except RFNU, 
RFNTL and RFNTS also perform well for a training ratio lower than 0.5. 
Overall, the accuracy of the methods is much higher than without bias: 
the file bias is very strong in this dataset, as illustrated by the large number
of constant lines in the matrices in Figure~\ref{fig:utility-matrices}.

\begin{figure*}
\begin{subfigure}[b]{0.4\linewidth}
        \includegraphics[width=0.8\columnwidth, angle=-90]{data/results/means_of_results/ALS-Bias/PreFreeSurfer/ALS-Bias-PFS-5v6.pdf}
        \caption{PFS, CentOS5 vs CentOS6}
\end{subfigure}
\begin{subfigure}[b]{0.4\linewidth}
        \includegraphics[width=0.8\columnwidth, angle=-90]{data/results/means_of_results/ALS-Bias/PreFreeSurfer/ALS-Bias-PFS-5v7.pdf}
        \caption{PFS, CentOS5 vs CentOS7}
        \label{fig:pfs-c5vsc7-bias}
\end{subfigure}
\begin{subfigure}[b]{0.4\linewidth}
        \includegraphics[width=0.8\columnwidth, angle=-90]{data/results/means_of_results/ALS-Bias/PreFreeSurfer/ALS-Bias-PFS-6v7.pdf}
        \caption{PFS, CentOS6 vs CentOS7}
\end{subfigure}\hfill
\begin{subfigure}[b]{0.4\linewidth}
        \includegraphics[width=0.8\columnwidth, angle=-90]{data/results/means_of_results/ALS-Bias/FS-100files/ALS-Bias-FS100files.pdf}
        \caption{FS, CentOS6 vs CentOS7}
\end{subfigure}
\caption{Accuracy results on PreFresurfer (PFS) and Freesurfer (FS) data, ALS \emph{with} bias.}
\label{fig:results-real-als-bias}
\end{figure*}

\todo{Different number of factors.}

\section{Discussion}

% Which method is the best
While collaborative filtering, perhaps unsurprisingly, is able to 
correctly predict the missing values in a matrix modeling the 
reproducibility problem, the usual random sampling method cannot be 
used in time-constrained processes. We proposed 6 sampling methods to 
address this issue, and we found that one of them, RFNU, (\todo{check 
that}) was consistently able to solve the problem. We explain that by 
the fact that RFNU leads the training set to contain a balanced mix 
between nearly complete and nearly empty columns, with a continuum of 
intermediate configurations in between. On the contrary, other sampling 
methods, including RFNT-L and RFNT-S, bias the sampling toward complete 
or empty columns, which is sometimes detrimental to 
accuracy.

% Training ratio
On datasets that are strongly dependent on row bias, such as the real 
datasets studied here, RFNU provides accuracy values that are 
consistently higher than 0.95 for training ratios higher than 0.5, even 
when biases are not included in the collaborative filtering 
optimization. On the synthetic dataset, RFNU still performs very well 
when biases are not included. However, the training ratio required to 
reach a 0.95 accuracy increases with the number of subject types 
present in the dataset. \todo{maybe find a relation between the 
required training ratio and the number of subjects per subject type in 
the entire dataset}.

% How can this be used in practice
From a practical standpoint, our study shows that reproducibility 
evaluations can be conducted using only 50\% of the files produced by 
the evaluated pipelines, with an accuracy above 95\%. Potentially, this 
could reduce the computing time and storage required for such studies 
by a factor of 2. However, such studies could not be conducted by 
processing only half of the subjects entirely, which would correspond 
to the Complete Columns sampling method. Instead, the processing of all 
the subjects should be initiated, and it should be terminated in a random
uniform way, assuming that files are uniformly produced throughout the 
execution.

% Generalizability

% What are the limitations
Our study could be extended by considering real-valued reproducibility 
matrices instead of just binary ones. It is indeed common for file 
differences to be quantified using specific similarity measures or 
distances, such as Levenshtein distance between strings, or the sum of 
squared distances between voxels of an image. Our sampling methods 
could be directly applied to real-valued matrices, and we expect our 
conclusions regarding the best-performing sampling method (RFNU) and 
the inclusion of biases in the model to remain valid.

% Real values, we expect the best sampling method to remain the same.
We recommend not to include biases in the collaborative filtering optimization
to solve this problem. Even though biases provide a slight accuracy improvement 
when datasets are strongly biased (constant lines in the matrix), they can also
be very detrimental for more complex datasets such as the syntetic one used here.
